Towards Practical First-Order Model Counting

### 2025-01-16 doc/paper/paper.tex
Rewrite Section 2.1

% TODO (for the entire paper):
% * 15 pages excluding references

@@

\section{Preliminaries}\label{sec:preliminaries}

% TODO: Section 2 until 2.3 (Algebra) could be drastically shortened

@@

% TODO: Skolemization is a procedure that transforms a formula with existential
% quantifiers into an ’equivalent’ formula without existential quantifiers with
% the same WFOMC.
In this section, we will review the basic concepts of first-order logic as they
are used in FOKC algorithms. There are two key differences between the logic
used by these algorithms and the logic supported as input. First,
Skolemization~\cite{DBLP:conf/kr/BroeckMD14} eliminates existential quantifiers
by introducing additional predicates. Please note that Skolemization here
differs from the standard Skolemization procedure that introduces function
symbols~\cite{DBLP:books/daglib/0030198}. Second, the input formula is rewritten
as a conjunction of clauses, each in \emph{prenex normal
  form}~\cite{hinman2018fundamentals}.

A \emph{term} can be either a variable or a constant. An \emph{atom} can be
either $P(t_{1}, \dots, t_{m})$ (or $P(\mathbf{t})$) for some predicate $P$ and
terms $t_{1}, \dots, t_{m}$ or $x=y$ for some terms $x$ and $y$. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of the predicate $P$ mentioned above. We write $P/m$ to denote a
predicate along with its arity. A \emph{literal} can be either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
}\forall x_{2} \in \Delta_{2}\dots\text{ }\forall x_{n} \in \Delta_{n}\text{.
}\phi(x_{1}, x_{2}, \dots, x_{n})$, where $\phi$ is a disjunction of literals
that only contain variables $x_{1}, \dots, x_{n}$ (and any constants). We say
that a clause is a \emph{(positive) unit clause} if there is only one literal
with a predicate, and it is a positive literal. Finally, a \emph{formula} is a
conjunction of clauses. Throughout the paper, we will use set-theoretic
notation, interpreting a formula as a set of clauses and a clause as a set of
literals.

% TODO: Page 2, Section 2.1, paragraph 2, line 10 - ”A clause is of the form ...” - are the quantifiers really part of the clause? Shouldn’t it be possible to quantify a conjunction of several clauses?
% Conforming with previous work, the definition of a clause includes universal quantifiers for all variables within. While it is possible to rewrite the entire formula in prenex normal form, the format we describe has proven itself convenient to work with.

@@

\begin{remark}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm can compile the formula into a
  circuit or graph suitable for lifted model counting. However, it is known that
  \textsc{ForcLift} compilation is guaranteed to succeed on any \FO{} formula
  without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
\end{remark}

@@

\begin{remark}
  The distinctness of domains is important in two ways. First, in terms of
  expressiveness, a clause such as $\forall x \in \Delta\text{. }P(x, x)$ is
  valid if predicate $P$ is defined over two copies of the same domain and
  invalid otherwise. Second, having more distinct domains makes the problem more
  decomposable for the FOKC algorithm. With distinct domains, the algorithm can
  make assumptions or deductions about, e.g., the first domain of predicate $P$
  without worrying how (or if) they apply to the second domain.
\end{remark}

@@

\begin{remark}
  Although \CompileWithBaseCases starts with a call to \textsc{Crane}, the
  proposed algorithm is not just a post-processing step for FOKC because
  \cref{alg:compilewithbasecases} is recursive and can issue more calls to
  \textsc{Crane} on various derived formulas.
\end{remark}

>>>>

% TODO (for the entire paper):
% * sentence vs formula: be consistent and not confusing
% * 15 pages excluding references

@@

\section{Preliminaries}\label{sec:preliminaries}
% TODO: focus more on current work
% TODO: Section 2 until 2.3 (Algebra) could be drastically shortened

@@

In this section, we will review the basic concepts of first-order logic as they
are used in FOKC algorithms. We begin by introducing the format used internally
by \textsc{ForcLift} and its descendants. Afterwards, we provide a high-level
description of how an arbitrary sentence in first-order logic is transformed
into this internal format.

A \emph{term} can be either a variable or a constant. An \emph{atom} can be
either $P(t_{1}, \dots, t_{m})$ (i.e., $P(\mathbf{t})$) for some predicate $P$
and terms $t_{1}, \dots, t_{m}$ or $x=y$ for some terms $x$ and $y$. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of the predicate $P$ mentioned above. We write $P/m$ to denote a
predicate along with its arity. A \emph{literal} can be either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom is \emph{ground} if it contains no variables, i.e., only constants. A
\emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
}\forall x_{2} \in \Delta_{2}\dots\text{ }\forall x_{n} \in \Delta_{n}\text{.
}\phi(x_{1}, x_{2}, \dots, x_{n})$, where $\phi$ is a disjunction of literals
that only contain variables $x_{1}, \dots, x_{n}$ (and any constants). We say
that a clause is a \emph{(positive) unit clause} if there is only one literal
with a predicate, and it is a positive literal. Finally, a \emph{formula} is a
conjunction of clauses. Throughout the paper, we will use set-theoretic
notation, interpreting a formula as a set of clauses and a clause as a set of
literals.

\begin{remark*}
  Conforming with previous work~\cite{DBLP:conf/ijcai/BroeckTMDR11}, the
  definition of a clause includes universal quantifiers for all variables
  within. While it is possible to rewrite the entire formula with all
  quantifiers at the front~\cite{hinman2018fundamentals}, the format we describe
  has proven itself convenient to work with.
\end{remark*}

There are two crucial preprocessing steps that transform an arbitrary sentence
in first-order logic into the form used internally: Skolemization and rewriting
the sentence as a conjunction of clauses. We describe the former in more detail.
\emph{Skolemization}~\cite{DBLP:conf/kr/BroeckMD14} is a procedure that
transforms a formula with existential quantifiers into a formula without
existential quantifiers \emph{with the same WFOMC}. (Note that it is different
from the standard Skolemization that introduces function
symbols~\cite{DBLP:books/daglib/0030198}.)

\begin{example}
  Skolemization transforms
  \begin{equation}\label[formula]{eq:skolemizationinitial}
    \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
  \end{equation}
  into
  \begin{equation}\label[formula]{eq:skolemization}
    \begin{aligned}
      (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
      (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)).
    \end{aligned}
  \end{equation}
  We will see how, with suitable weights on the new predicates $S/1$ and $Z/1$,
  the WFOMC of \cref{eq:skolemization} is equal to the FOMC of
  \cref{eq:skolemizationinitial}.
\end{example}
% TODO: make sure to revisit this in the future

@@

\begin{remark*}
  In the case of \textsc{ForcLift} and its extensions, support for a formula as
  valid input does not imply that the algorithm can compile the formula into a
  circuit or graph suitable for lifted model counting. However, it is known that
  \textsc{ForcLift} compilation is guaranteed to succeed on any \FO{} formula
  without constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
\end{remark*}

@@

\begin{remark*}
  The distinctness of domains is important in two ways. First, in terms of
  expressiveness, a clause such as $\forall x \in \Delta\text{. }P(x, x)$ is
  valid if predicate $P$ is defined over two copies of the same domain and
  invalid otherwise. Second, having more distinct domains makes the problem more
  decomposable for the FOKC algorithm. With distinct domains, the algorithm can
  make assumptions or deductions about, e.g., the first domain of predicate $P$
  without worrying how (or if) they apply to the second domain.
\end{remark*}

@@

\begin{remark*}
  Although \CompileWithBaseCases starts with a call to \textsc{Crane}, the
  proposed algorithm is not just a post-processing step for FOKC because
  \cref{alg:compilewithbasecases} is recursive and can issue more calls to
  \textsc{Crane} on various derived formulas.
\end{remark*}
